---
title: "awu3_hw2"
format: html
editor: visual
---

### Task 1: Basic Vector Practice

-   Goal: Create 2 basic vectors of blood pressure patient before and after treatment and do some basic functions on the vector

```{r}

# (1) create vector with pre-treatment values
pre_bp <- c(130, 128, 116, 124, 133, 
            134, 118, 126, 114, 127, 
            141, 138, 128, 140, 137, 
            131, 120, 128, 139, 135)

# (1) created vector with post-treatment values
post_bp <- c(114, 98, 113, 99, 107, 116,
             113, 111, 119, 117, 101, 119, 
             130, 122, 106, 106, 124, 102, 
             117, 113)

# (2) assign names
names(pre_bp) <- paste("Subject", 1:20, sep = "_")

names(post_bp) <- paste("Subject", 1:20, sep = "_")

# (3) vector with post blood pressure - pre blood pressure
diff_bp <- c(post_bp - pre_bp)

diff_bp

# (4) average diff in blood pressure
mean(diff_bp)

# (5) find patients who had a decrease in blood pressure after treatment
which(diff_bp < 0)

# (6) Only return(subset of the vector) positive change in the difference vector
x <- c(9, 13, 17)
diff_bp[-x]


# (7) calculate average decrease in blood pressure from those with a positive change
mean(diff_bp[-x])


```

### Task 2: Basic Data Frame Practice

-   Goal: From the previous example create a new dataframe and use some conditional logic to return certain rows and create a new column

```{r}

# make patient names vector
patient <- c(names(pre_bp))

# (1) create data frame with 4 cols patient, pre_bp, post_bp, and diff_bp
bp_df <- data.frame(patient, pre_bp, post_bp, diff_bp,row.names= NULL )

bp_df

# (2) returns rows where diff_bp is negative
subset(bp_df, diff_bp < 0)

# (3) create column where TRUE if post bp is less than 120 

bp_df$optimal_bp <- bp_df$post_bp < 120

# (4) print data frame 

knitr::kable(bp_df)

```

### Task 3: List Practice

-   Goal: Create a list of dataframes and access the elements in the list

```{r}
# (1) create new vectors for placebo treatments df

pre_bp_placebo <- c(138, 135, 147, 117, 152, 134, 114, 121, 131, 130)

post_bp_placebo <- c(105, 136, 123, 130, 134, 143, 135, 139, 120, 124)

diff_bp_placebo <- post_bp_placebo - pre_bp_placebo

patient_placebo <- paste("Subject", 1:10, sep = "_")
  
# (1) create new data frame for placebo patients
placebo_df <- data.frame(patient_placebo,pre_bp_placebo, post_bp_placebo,diff_bp_placebo)

# create column for post bp <120

placebo_df$optimal_bp<- placebo_df$post_bp_placebo < 120

placebo_df

# (2) create list with 2 df elements created above

df_list <- list(treatment = bp_df, placebo = placebo_df)

# (3) access df_list[1] multiple ways
df_list[[1]]

df_list$treatment

df_list[1]

# (4) access placebo df pre_bp column
df_list$placebo$pre_bp_placebo
```

### Task 4: Control Flow Practice

-   Goal: From the previous example create a new status column using for loops with conditional logic to label rows

```{r}
# (1)create new column status for both data frames
df_list$treatment$status <- character(20)

df_list$placebo$status <- character(10)

# (2) create status column values for status in treatment df
for (i in seq_along(df_list$treatment$post_bp)) {
  if (df_list$treatment$post_bp[i] <= 120) {
    df_list$treatment$status[i] = 'optimal'
  }
  else if (df_list$treatment$post_bp[i] <= 130) {
    df_list$treatment$status[i] = 'boderline'
  }
  else {
    df_list$treatment$status[i] = 'high'
  }
} 

df_list$treatment

# (3) create status column values for status in placebo df
for (i in seq_along(df_list$placebo$post_bp_placebo)) {
  if (df_list$placebo$post_bp_placebo[i] <= 120) {
    df_list$placebo$status[i] = 'optimal'
  }
  else if (df_list$placebo$post_bp_placebo[i] <= 130) {
    df_list$placebo$status[i] = 'boderline'
  }
  else {
    df_list$placebo$status[i] = 'high'
  }
} 

df_list$placebo

```
