---
title: "bdevapa_hw2.qmd"
format: html
editor: visual
---

# Task 1: Basic Vector Practice

## 1.q Pre-treatment blood pressure

```{r}
pre_treatment_bp <- c(130, 128, 116, 124, 133, 134, 118, 126, 114, 127, 141, 138, 128, 140, 137, 131, 120, 128, 139, 135)
```

## 1.b Post-treatment blood pressure

```{r}
post_treatment_bp <- c(114, 98, 113, 99, 107, 116, 113, 111, 119, 117, 101, 119, 130, 122, 106, 106, 124, 102, 117, 113)
```

## 2. Assign names to both vectors

```{r}
# create names once and then reuse them 
subject_names <- paste("Subject", 1:20, sep = "-")
names(pre_treatment_bp) <- c(subject_names)
names(post_treatment_bp) <- c(subject_names)
pre_treatment_bp
pre_treatment_bp
```

## 3. Change in blood pressure after treatment

```{r}
change_in_bp <- pre_treatment_bp - post_treatment_bp
change_in_bp
```

## 4. Average decrease in blood pressure accorss all patients

```{r}
avg_decrease_in_bp <- mean(change_in_bp)
avg_decrease_in_bp
```

## 5. Patients experiencing positive change after treatment

```{r}
patients_with_positive_change <- which(change_in_bp > 0)
patients_with_positive_change
```

## 6. Subset of the vector of differences with positive change after treatment

```{r}
positive_difference_in_bp <- change_in_bp[change_in_bp > 0]
positive_difference_in_bp
```

## 7. Average of positive differences after treatment after treatment

```{r}
avg_positive_differennce_in_bp <- mean(positive_difference_in_bp)
avg_positive_differennce_in_bp
```

# Task 2: Basic Data Frame practice

## 1. Create blood pressure data frame

```{r}
treatment_df <- data.frame(
  patient = subject_names,
  pre_bp = pre_treatment_bp,
  post_bp = post_treatment_bp,
  diff_bp = change_in_bp
)

str(treatment_df)
```

## 2. Patients with increased blood pressure after treatment

```{r}
patients_with_increased_blood_pressure_df <- treatment_df[treatment_df$diff_bp < 0, ]

patients_with_increased_blood_pressure_df
```

## 3. Add new column to label healthy blood pressure post treatment

```{r}
treatment_df$healthy_bp_post_treatment <- treatment_df$post_bp < 120
treatment_df
```

## 4. Print data frame

```{r}
knitr::kable(treatment_df)
```

# Task 3: List practice

## 1. Placebo treatment vectors and then data frame

```{r}
pre_placebo_bp <- c(138, 135, 147, 117, 152, 134, 114, 121, 131, 130)
post_placebo_bp <- c(105, 136, 123, 130, 134, 143, 135, 139, 120, 124)
placebo_subject_names <- paste("Subject", 21:30, sep = "-")
names(pre_placebo_bp) <- c(placebo_subject_names)
names(pre_placebo_bp) <- c(placebo_subject_names)
placebo_df <- data.frame(
  pre_bp = pre_placebo_bp,
  post_bp = post_placebo_bp
)

str(placebo_df)
```

## 2. Create list with data frames

```{r}
treatment_list <- list(
  treatment = treatment_df,
  placebo = placebo_df
)
```

## 3. Access the first element with 3 different syntax

```{r}
treatment_list$treatment
treatment_list[1]
treatment_list[[1]]

```

## 3. In one line access placebo data frame, pre_bp column

```{r}
treatment_list$placebo$pre_bp

```

# Task 4: Control Flow Practice

## 1. Add new column to characterize post treatment blood flow as optimal (≤ 120), borderline (120 \< bp ≤ 130), and high (\> 130)

```{r}
# instead of hardcoding use nrow to find the row dimension of the data frame 
treatment_df$status <- character(nrow(treatment_df))
str(treatment_df)

placebo_df$status <- character(nrow(placebo_df))
str(placebo_df)


```

## 2. Assign status column to blood_pressure_df (non-placebo data frame)

```{r}
for(i in 1:nrow(treatment_df)) {
  # print(i)
  # print(blood_pressure_df$post_bp[i])
  if(treatment_df$post_bp[i] <= 120) {
     treatment_df$status[i] <- 'optimal'
  } else if(treatment_df$post_bp[i] <= 130) {
    treatment_df$status[i] <- 'borderline'
  } else if(treatment_df$post_bp[i] > 130) {
    treatment_df$status[i] <- 'high'
  }
}

treatment_df

```

## 3. Assign status column to blood_pressure_df (non-placebo data frame)

```{r}
for(i in 1:nrow(placebo_df)) {
  # print(i)
  # print(blood_pressure_df$post_bp[i])
  if(placebo_df$post_bp[i] <= 120) {
     placebo_df$status[i] <- 'optimal'
  } else if(placebo_df$post_bp[i] <= 130) {
    placebo_df$status[i] <- 'borderline'
  } else if(placebo_df$post_bp[i] > 130) {
    placebo_df$status[i] <- 'high'
  }
}

placebo_df

```
