---
title: "ST558 Homework 2"
author: "Lasya Harini Banka"
format: html
editor: visual
---

**Task 1: Basic Vector practice**

Suppose we have data from a medical experiment on blood pressure. We have the following pre-treatment values for subjects 1 through 20:

• 130, 128, 116, 124, 133, 134, 118, 126, 114, 127, 141, 138, 128, 140, 137, 131, 120, 128, 139, 135

after treatment, the subjects were measured again (subjects 1 through 20 match)

• 114, 98, 113, 99, 107, 116, 113, 111, 119, 117, 101, 119, 130, 122, 106, 106, 124, 102, 117, 113

**1)** Create two vectors. One vector corresponding to the pre measurements and one to the post measurements.

**Solution:**

```{r}
# Pre-treatment blood pressure measurements
pre_bp <- c(
  130, 128, 116, 124, 133,
  134, 118, 126, 114, 127,
  141, 138, 128, 140, 137,
  131, 120, 128, 139, 135
)

# Post-treatment blood pressure measurements
post_bp <- c(
  114, 98, 113, 99, 107,
  116, 113, 111, 119, 117,
  101, 119, 130, 122, 106,
  106, 124, 102, 117, 113
)

pre_bp
post_bp
```

Above I created one vector for the pre-treatment measurements and another vector for the post-treatment measurements.

**2)** Assign names to the vector elements using the paste() function. Note that names() can be overwritten by a character vector. To quickly create the names, try running the code paste. Create the same names for each vector’s elements

**Solution:**

I assigned same subject names to both vectors below to make it easy to identify the measurements for each subject.

```{r}
#assigning names to the above created vector elements
subject_names <- paste("Subject", 1:20, sep = "_")

names(pre_bp) <- subject_names
names(post_bp) <- subject_names

pre_bp
post_bp
```

**3)** Calculate the change in blood pressure for each patient by subtracting post-treatment measurements from pre-treatment measurements. Recall that R does math element-wise! Save this calculation as a new object (also a vector).

**Solution:**

```{r}
# Calculate the change in blood pressure
change_bp <- pre_bp - post_bp

change_bp
```

A positive value means that the blood pressure decreased after treatment and its the opposite for a negative value (increased blood pressure).

**4)** Calculate the average decrease in blood pressure across all patients.

**Solution:**

I used the mean() function to calculate the average change across all 20 subjects.

```{r}
# Calculate the average change for all subjects
mean(change_bp)
```

From the result above we note that the average change is 17.

**5)** Determine which patients experienced a decrease in blood pressure after treatment (a positive change). Use the which() function to just return the indices (and names) associated with this type of change.

**Solution:**

A positive change means that the pre-treatment blood pressure was greater than the post-treatment blood pressure. I used the which() to find the positions where the change is greater than zero.

```{r}
# Find the subjects with a positive change
which(change_bp > 0)
```

There are 17 subjects with a positive change. However Subjects 9, 13 and 17 are not in the list above since their blood pressure went up.

**6)** Subset the vector of differences to only return those that have a positive change.

**Solution:**

Now I subset the change_bp vector so that only positive changes are returned.

```{r}
# Keeping only the positive changes
positive_changes <- change_bp[change_bp > 0]

positive_changes
```

From above output we note that the subjects 9, 13 and 17 had increased blood pressure after treatment.

**7)** Calculate the average decrease in blood pressure for those where the blood pressure decreased (positive change).

**Solution:**

```{r}
# Calculate the average decrease among subjects whose blood pressure decreased
mean(positive_changes)
```

The average decrease among the subjects whose blood pressure decreased is 20.647.

**Task 2: Basic Data Frame practice**

**1)** Create a data frame object with four columns corresponding to your data above: patient, pre_bp, post_bp, and diff_bp

**Solution:**

```{r}
#creating blood pressure dataframe
bp_df <- data.frame(
  patient = names(pre_bp),
  pre_bp=pre_bp,
  post_bp=post_bp,
  diff_bp=change_bp
)
bp_df
```

Dataframe created successfully with four columns as shown above.

**2)** Return only rows where the diff_bp column is negative. (Use \[ or learn about the subset() function if you’d like. If you use \[, don’t reference the original vector from the first part, access the column of the data frame when looking at making a comparison.)

**Solution:**

```{r}
#Return rows where blood pressure change is negavtive
bp_df[bp_df$diff_bp<0, ]
```

These patients are the same numbers that had increased blood pressure after treatment .

**3)** Add a new column to the data frame corresponding to TRUE if the post_bp is less than 120. Recall you can use \$ to reference a column. If you reference a column that doesn’t exist, and save a vector (of appropriate length in it), that vector becomes a column of your data frame! Similar to the previous question, don’t reference the original vector from the first part, access the column of the data frame when looking at making a comparison.

**Solution:**

```{r}
#Add True or False based on post treatment blood pressure
bp_df$post_under_120 <- bp_df$post_bp <120
bp_df
```

If post treatment blood pressure is less than 120 then its value us 'True', else its 'False'.

**4)** Finally, print the data frame out nicely in your final document by modifying the code below appropriately.

knitr::kable(bp_df)

**Solution:**

```{r}
#Above table displayed as dataframe
knitr::kable(bp_df)
```

**Task 3: List practice**

Continue the previous example. Suppose we now also have data from another experiment where the ‘treat- ment’ was actually a placebo.

We have the following pre-treatment values for subjects 1 through 10 (different set of subjects):

• 138, 135, 147, 117, 152, 134, 114, 121, 131, 130

after treatment, the subjects were measured again (subjects 1 through 10 match)

• 105, 136, 123, 130, 134, 143, 135, 139, 120, 124

**1)** Create a new data frame with this data in it that is similar to the data frame from task 2 (including the new column)

**Solution:**

I first created vectors for the placebo groups pre and post treatment blood pressure measurement. Then calculated the differences between them to create a data frame similar to the dataframe for treatment above.

```{r}
#pre treatment blood pressure measurements for the placebo group
placebo_pre <- c(138, 135, 147, 117, 152, 134, 114, 121, 131, 130)
#post treatment blood pressure measurements for the placebo group
placebo_post <- c(105, 136, 123, 130, 134, 143, 135, 139, 120, 124)

#calculate the chances in blood pressure
placebo_diff <- placebo_pre - placebo_post

#get the subject names
placebo_names <- paste("Subject", 1:10, sep="_")

#Placebo dataframe creation
placebo_df <- data.frame(
  patient = placebo_names,
  pre_bp = placebo_pre,
  post_bp = placebo_post,
  diff_bp = placebo_diff
)

#adding a column showing whether post treatment blood pressure is below 120
placebo_df$post_under_120 <- placebo_df$post_bp <120

placebo_df
```

**2)** Now create and store a list with two elements:

• 1st element named treatment and contains the first data frame you created.

• 2nd element named placebo and contains the second data frame you created.

**Solution:**

```{r}
#two lists creation done
bp_list <- list( treatment =bp_df,
                 placebo = placebo_df)

bp_list
```

**3)** Access the first list element using three different types of syntax.

**Solution:**

Accessing treatment data using three different types of status as below.

```{r}
#treatment dataframe using name
bp_list[["treatment"]]

#treatment dataframe using position
bp_list[[1]]

#treatment dataframe using $
bp_list$treatment
```

**4)** In one line, access the placebo data frame, pre_bp column.

**Solution:**

I accessed the 'pre_bp' column from the placebo data frame from the list using '\$' syntax.

```{r}
#Access to the placebo pre treatment bloodpressure column
bp_list$placebo$pre_bp
```

**Task 4: Control Flow Practice**

**1)** Suppose we want to characterize the post-treatment (or placebo) blood pressure measurement as optimal (≤ 120), borderline (120 \< bp ≤ 130), and high (\> 130). First, create a new column in each data frame from above called status.

You can do this via:

your_df\$status \<- character(20) #or 10 depending on number of observations

Note: You want to add this additional column to the data frames that are stored in your list (R doesn’t do referencing to the original object).

**Solution:**

I add a empty character column called status for both the data frames stores in the list. Also note that treatment data has 20 observations and placebo data has 10 observations.

```{r}
#empty status column for treatment data
bp_list$treatment$status <- character(20)

#empty empty status column for placebo data
bp_list$placebo$status <- character(10)

bp_list$treatment
bp_list$placebo
```

**2)** For the non-placebo data frame (within the list), create a for loop and use if/then/else logic to create the status column’s values.

**Solution:**

Created a for loop (through 20 subjects) below to classify the treatment blood pressure into the 3 stages - 'optimal', 'borderline' and 'high'. As mentioned above, the post-treatment blood pressure measurement as optimal (\<= 120), borderline (120 \< bp \<= 130), and high (\> 130). Same logic is applied using if, else and else if in the code below.

```{r}
#Using for loop to classify treatment blood pressure into 3 stages 
for (i in 1:20){
  if (bp_list$treatment$post_bp[i] <= 120){
    bp_list$treatment$status[i] <- "optimal"
  }
  else if(bp_list$treatment$post_bp[i] <= 130){
    bp_list$treatment$status[i] <- "borderline"
  } else {
    bp_list$treatment$status[i] <- "high"
  }
}

bp_list$treatment
```

**3)** Repeat for the placebo data frame (within the list). • In the final document, make sure that all code chunks show (and are evaluated). Use headings to separate the sections. Write text before each code chunk explaining what you are trying to do. Use markdown where appropriate (to create lists, bold things, etc.).

Include some basic comments in the code where needed.

**Solution:**

Same process as above is repeated, but for placebo data with 10 subjects. Similarly as mentioned above the placebo blood pressure measurement as optimal (\<=120), borderline (120 \< bp \<= 130), and high (\> 130). Same logic is applied using if, else and else if in the code below.

```{r}
#Using for loop to classify placebo blood pressure into 3 stages 
for(i in 1:10){
  if (bp_list$placebo$post_bp[i] <= 120){
    bp_list$placebo$status[i] <- "optimal"
  }
  else if(bp_list$placebo$post_bp[i] <= 130){
    bp_list$placebo$status[i] <- "borderline"
  } else {
    bp_list$placebo$status[i] <- "high"
  }
}

bp_list$placebo

```
