---
title: "ST558 Homework 2"
author: "Lasya Harini Banka"
format: html
editor: visual
---

**Task 1: Basic Vector practice**

Suppose we have data from a medical experiment on blood pressure. We have the following pre-treatment values for subjects 1 through 20:

• 130, 128, 116, 124, 133, 134, 118, 126, 114, 127, 141, 138, 128, 140, 137, 131, 120, 128, 139, 135

after treatment, the subjects were measured again (subjects 1 through 20 match)

• 114, 98, 113, 99, 107, 116, 113, 111, 119, 117, 101, 119, 130, 122, 106, 106, 124, 102, 117, 113

**1)** Create two vectors. One vector corresponding to the pre measurements and one to the post measurements.

**Solution:**

```{r}
# Pre-treatment blood pressure measurements
pre_bp <- c(
  130, 128, 116, 124, 133,
  134, 118, 126, 114, 127,
  141, 138, 128, 140, 137,
  131, 120, 128, 139, 135
)

# Post-treatment blood pressure measurements
post_bp <- c(
  114, 98, 113, 99, 107,
  116, 113, 111, 119, 117,
  101, 119, 130, 122, 106,
  106, 124, 102, 117, 113
)

pre_bp
post_bp
```

Above I created one vector for the pre-treatment measurements and another vector for the post-treatment measurements.

**2)** Assign names to the vector elements using the paste() function. Note that names() can be overwritten by a character vector. To quickly create the names, try running the code paste. Create the same names for each vector’s elements

**Solution:**

```{r}
subject_names <- paste("Subject", 1:20, sep = "_")

names(pre_bp) <- subject_names
names(post_bp) <- subject_names

pre_bp
post_bp
```

**3)** Calculate the change in blood pressure for each patient by subtracting post-treatment measurements from pre-treatment measurements. Recall that R does math element-wise! Save this calculation as a new object (also a vector).

**Solution:**

```{r}
# Calculate the change in blood pressure
change_bp <- pre_bp - post_bp

change_bp
```

A positive value means that the blood pressure decreased after treatment.

**4)** Calculate the average decrease in blood pressure across all patients.

**Solution:**

I used the mean() function to calculate the average change across all 20 subjects.

```{r}
# Calculate the average change for all subjects
mean(change_bp)
```

From the result above we note that the average change is 17 mmHg.

**5)** Determine which patients experienced a decrease in blood pressure after treatment (a positive change). Use the which() function to just return the indices (and names) associated with this type of change.

**Solution:**

A positive change means that the pre-treatment blood pressure was greater than the post-treatment blood pressure. I used the which() to find the positions where the change is greater than zero.

```{r}
# Find the subjects with a positive change
which(change_bp > 0)
```

There are 17 subjects with a positive change. However Subjects 9, 13 and 17 are not in the list above since their blood pressure went up.

**6)** Subset the vector of differences to only return those that have a positive change.

**Solution:**

Now I subset the change_bp vector so that only positive changes are returned.

```{r}
# Keeping only the positive changes
positive_changes <- change_bp[change_bp > 0]

positive_changes
```

From above output we note that the subjects 9, 13 and 17 had increased blood pressure after treatment .

**7)** Calculate the average decrease in blood pressure for those where the blood pressure decreased (positive change).

**Solution:**

```{r}
# Calculate the average decrease among subjects whose blood pressure decreased
mean(positive_changes)
```

The average decrease among the subjects whose blood pressure decreased is 20.647.
